Implementing Distributed Systems Directly with I/O Automata in Ioa++

### 2011-09-16 papers/paper1/programming_model.tex
Case study.

\subsection{Dynamic Composition}

\begin{figure}
\center
\includegraphics[width=\textwidth]{tcp_lcr_automaton}
\caption{TCP LCR automaton.}
\label{tcp_lcr_automaton}
\end{figure}

\paragraph{Leader election in a ring using TCP.}
The \verb+asynch_lcr_automaton+ is a reusable component designed to be used in the context of a larger system.
The \verb+tcp_lcr_automaton+ listed in \verb+examples/tcp_lcr.cpp+ elects a leader in a dynamic ring of processes connected by TCP sockets.
The structure of the \verb+tcp_lcr_automaton+ is given in Figure~\ref{tcp_lcr_automaton}.
The \verb+tcp_lcr_automaton+ is composed with a \verb+tcp_acceptor_automaton+ to accept a connection from its predecessor, a \verb+tcp_connector_automaton+ to create a connection to its successor, two \verb+tcp_connection_automaton+ representing the predecessor and successor, and a \verb+asynch_lcr_automaton+ to realize the leader election algorithm.
The leader of the ring periodically sends a token which is forwarded by all nodes in the ring.
If a node has not received a token for a specified duration of time, it calls for an election.

\paragraph{Binding predicates.}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the \verb+ioa::binding_count+ function which returns the number of actions bound to the specified action.
The value returned by \verb+ioa::binding_count+ is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, \verb+ioa::binding_count+ is used in the precondition of outputs to ensure that the value generated by the output will be received.
This type of usage can be seen in the send action of the \verb+tcp_lcr_automaton+:
\begin{lstlisting}
  bool send_precondition (
    ioa::automaton_manager<ioa::tcp_connection_automaton>* s) const
  {
    return ... && ioa::binding_count (&tcp_lcr_automaton::send, s) != 0;
  }
\end{lstlisting}

\paragraph{Dynamic composition.}
Dynamic composition is accomplished by creating child automata and bindings via \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+ objects.
The \verb+tcp_lcr_automaton+ creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
\begin{lstlisting}
    ioa::automaton_manager<asynch_lcr_automaton<uuid> >* lcr =
       ioa::make_automaton_manager (this,
         ioa::make_generator<asynch_lcr_automaton<uuid> > (m_u));
    ioa::make_binding_manager (this,
			       lcr, &asynch_lcr_automaton<uuid>::send,
			       &m_self, &tcp_lcr_automaton::send_lcr);
    ioa::make_binding_manager (this,
			       &m_self, &tcp_lcr_automaton::receive_lcr,
			       lcr, &asynch_lcr_automaton<uuid>::receive);
    ioa::make_binding_manager (this,
			       lcr, &asynch_lcr_automaton<uuid>::leader,
			       &m_self, &tcp_lcr_automaton::leader_lcr);
    ioa::make_binding_manager (this,
			       &m_self, &tcp_lcr_automaton::init_lcr,
			       lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
Children automata are managed by \verb+ioa::automaton_manager+ objects which can be created via the \verb+ioa::make_automaton_manager+ function.
An automaton manager object requires an \verb+ioa::automaton+ object for its system action implementation and a generator for allocating the child automaton.
Bindings are managed by \verb+ioa::binding_manager+ objects which can created via the \verb+ioa::make_binding_manager+ function.
A binding manager objects requires an \verb+ioa::automaton+ object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The \verb+ioa::automaton_manager+ class and \verb+ioa::binding_manager+ class contain methods for destroying and unbinding respectively.
The \verb+m_self+ object of the example is a \verb+ioa::handle_manager+ which can be used to bind to automata that already exist.
From the example, the \verb+tcp_lcr_automaton+ binds the \verb+asynch_lcr_automaton+ to itself.
The wary reader might notice that the \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+s in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the \verb+ioa::automaton+ and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.

@@

\paragraph{Dynamics in the TCP LCR automaton.}
A \verb+tcp_lcr_automaton+ continuously attempts to complete the ring by accepting a connection from its predecessor or connecting to its successor.
Accepting can fail if the socket used for listening is currently unavailable.
Connecting can fail if the successor automaton does not exist and is therefore not listening for incoming connections.
The connections themselves will fail when nodes leave the ring.
The \verb+tcp_lcr_automaton+ also contains constraints on the order in which automata are created with respect to each other and with respect to which actions are executed.
For example, the predecessor connection must be created before it is passed to the acceptor via the accept action.
Similarly, the successor connection must be created before it is passed via a constructor to the connector.

\paragraph{System action events.}
Determining the outcome of a system action event, e.g., that a child automaton was created, is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata supply the output action and input action are created before attempting to bind.
User automata can observe \verb+ioa::automaton_manager+ objects and \verb+ioa::binding_manager+ objects to receive events generated upon the completion of a system action.

@@

%% Conclude with our recommendations for programming, highlight need for exploration.

%% Tips for Writing Programs with I/O Automata
%% 1. Make sure that a parameter exists.
%% 2. Move logic in observe to local actions.
%% 3. Avoid observing an arbitrary number of managers.
%% 4. Do not bind to managers allocated on the stack.
%% 5. Stop when an error is encoutered and report it.
%% 6. Make sure output actions are bound.
%% 7. Make sure local actions are scheduled.
%% 8. Do not pass pointers---use const_shared_ptr.
%% 9. Check preconditions.
%% 10. Check effects.
%% 11. Order clauses of preconditions for efficiency.
%% 12. Automata that should be destroyed at the same time should be created at the same time and vice versa.  (Group using a parent.)

%% \begin{outline}
%% \item State
%%   \begin{outline}
%%   \item There is no shared state in a 
%%   \item Local state only
%%   \item Shared state
%%     \begin{outline}
%%       \item Impossible in distributed systems
%%       \item Dangerous in local systems
%%     \end{outline}
%%   \end{outline}
%% \item Communication
%%   \begin{outline}
%%   \item Atomic asynchronous message passing
%%   \item Network sets size of atom (UDP)
%%   \item Can build reliable streams (TCP)
%%   \item Local equivalent is passing a value
%%   \item Model should lend itself to writing protocols
%%   \end{outline}
%% \item Asynchrony
%%   \begin{outline}
%%     \item Model must have natural support for asynchrony, i.e., event-based
%%     \item Leads to a more efficient implementation because changed state and enabled actions become obvious
%%   \end{outline}
%% \item Concurrency
%%   \begin{outline}
%%     \item Reason about systems using non-deterministic interleaving of atomic actions
%%     \item Model should admit implementations that execute concurrently
%%   \end{outline}
%% \item Dynamics
%%   \begin{outline}
%%     \item Configuration - Edges in graph of communicating components can change at run-time.
%%       \begin{outline}
%%       \item Already required in distributed settings
%%       \item Not addressed in formal models
%%       \end{outline}
%%     \item Extension - Nodes in graph of communicating components can change at run-time.
%%   \end{outline}
%%   \item Reflection
%% \end{outline}

%% I/O Automata
%% \begin{itemize}
%%   \item Compare with UNITY
%%   \item Compare with esterel
%%   \item Compare with pi calculus
%%   \item Compare with Ptolemy
%% \end{itemize}

>>>>

\subsection{Dynamics}

\paragraph{Dynamic composition.}
The \verb+asynch_lcr_automaton+ is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+ objects.
The \verb+tcp_lcr_automaton+ discussed in section~\ref{case_study} creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
\begin{lstlisting}
    ioa::automaton_manager<asynch_lcr_automaton<uuid> >* lcr =
       ioa::make_automaton_manager (this,
         ioa::make_generator<asynch_lcr_automaton<uuid> > (m_u));
    ioa::make_binding_manager (this,
			       lcr, &asynch_lcr_automaton<uuid>::send,
			       &m_self, &tcp_lcr_automaton::send_lcr);
    ioa::make_binding_manager (this,
			       &m_self, &tcp_lcr_automaton::receive_lcr,
			       lcr, &asynch_lcr_automaton<uuid>::receive);
    ioa::make_binding_manager (this,
			       lcr, &asynch_lcr_automaton<uuid>::leader,
			       &m_self, &tcp_lcr_automaton::leader_lcr);
    ioa::make_binding_manager (this,
			       &m_self, &tcp_lcr_automaton::init_lcr,
			       lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
Children automata are managed by \verb+ioa::automaton_manager+ objects which can be created via the \verb+ioa::make_automaton_manager+ function.
An automaton manager object requires an \verb+ioa::automaton+ object for its system action implementation and a generator for allocating the child automaton.
Bindings are managed by \verb+ioa::binding_manager+ objects which can created via the \verb+ioa::make_binding_manager+ function.
A binding manager objects requires an \verb+ioa::automaton+ object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The \verb+ioa::automaton_manager+ class and \verb+ioa::binding_manager+ class contain methods for destroying and unbinding respectively.
The \verb+m_self+ object of the example is a \verb+ioa::handle_manager+ which can be used to bind to automata that already exist.
From the example, the \verb+tcp_lcr_automaton+ binds the \verb+asynch_lcr_automaton+ to itself.
The wary reader might notice that the \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+s in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the \verb+ioa::automaton+ and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.

\paragraph{Binding predicates.}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the \verb+ioa::binding_count+ function which returns the number of actions bound to the specified action.
The value returned by \verb+ioa::binding_count+ is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, \verb+ioa::binding_count+ is used in the precondition of outputs to ensure that the value generated by the output will be received.
This type of usage can be seen in the send action of the \verb+asynch_lcr_automaton+:
\begin{lstlisting}
  bool send_precondition () const {
    return !m_send.empty () &&
      ioa::binding_count (&asynch_lcr_automaton::send) != 0;
  }
\end{lstlisting}

@@

\paragraph{System action events.}
Determining the outcome of a system action event, e.g., that a child automaton was created, is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata supply the output action and input action are created before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage and therefore expects the automaton to exist before attempting to bind.
To enable such behavior, user automata can observe \verb+ioa::automaton_manager+ objects and \verb+ioa::binding_manager+ objects to receive events generated upon the completion of a system action.

@@
